---
title: "Analysis of Fares and Operational Efficiency"
output: rmarkdown::github_document
---
```{r setup, include=FALSE}
library(tidyverse)
```

We begin by pulling in fare data and specifying major airlines to analyze.

```{r}
flight_data <- read.csv('Data/PRIMARY_flight_traffic.csv')
fares <- read.csv('Data/fares.csv')

airline_ids_keep <- c("UA", "AA", "US", "F9", "B6", "OO", "AS", "NK", "WN", "DL", "EV", "HA", "MQ", "VX")

fares <- fares %>%
  filter(airline_id %in% airline_ids_keep)

fares[1,]
```

Right now, we don't have actual average fare data for the flight - we have the number of people who paid within a bucket (of size 10). We need to convert this to an average fare paid, per quarter, per route. 

```{r}

#calculate the price from the bucket data in the fares DB 
price <- seq(from = 15, to = 2505, by = 10) #assume in each bucket, avg fare is middle bucket

#get the total sum spent per bucket
sum_fares_per_bucket <- data.frame(mapply(`*`,fares[, 6:255],price))

#sum all the bucket fares and divide by the total number of people who bought tickets
totalFaresPerFlight <- rowSums(sum_fares_per_bucket)
totalPeoplePerFlight <- rowSums(fares[, 6:255])

fares$avg_fare <-  totalFaresPerFlight/totalPeoplePerFlight

select(fares, c("airline_id", "origin_airport", "destination_airport", "quarter", "avg_fare"))[1:5,]
```

Now that we have the average fare in USD paid per route in each quarter, we can aggregate this across quarters and split the data to look at fares by airline and route.

```{r}
#so i can join back in when I group by airline
route_distances <- fares %>%
  mutate(route = paste(origin_airport, destination_airport)) %>%
  select(c("route", "distance", "origin_airport", "destination_airport"))

#summarise the average fare per airline per route
route_fares <- fares %>%
  mutate(route = paste(origin_airport, destination_airport)) %>%
  group_by(airline_id, route, origin_airport, distance) %>%
  summarise(avg_fare = mean(avg_fare))

route_fares[1:5,]
```

The question we want to examine is: Holding the airline and route distance constant, do higher delays caused at airports correlate with lower fares at that airport? 

```{r}
#dataset created separately to group airlines with delays
delays <- read.csv('Data/delays.csv')

#Join route delays to fare information
route_fares_delays <- inner_join(route_fares, delays, by = 'origin_airport') %>%
  mutate(route_dist = cut(distance, breaks=seq(0,1500, by=100)))

#Filter to a specific airline for analysis
airline_filter <- "DL"

route_dist_by_airline <- route_fares_delays %>%
  group_by(airline_id, route_dist, origin_airport) %>%
  summarize(avg_fare_airport = mean(avg_fare), avg_delay_airport = mean(Avg..Total.Departure.Delay)) %>%
  filter(airline_id == "DL")

#hold distance constant, plot a scatterplot for each distance segment
unique_route <- unique(route_dist_by_airline$route_dist)

for(route in unique_route)
{
  selected_route <- filter(route_dist_by_airline, route_dist == route)
    
  plot(selected_route$avg_delay_airport, selected_route$avg_fare_airport, sub = paste("Flight Distance Between: ", route), main = paste(airline_filter, "Flights"), xlab="Delay (minutes)", ylab="Fare (USD $)")
      
  abline(lm(selected_route$avg_fare_airport ~ selected_route$avg_delay_airport))
}
```

Looking at the scatterplots, we have identified no clear trend between fare prices and the delays inherent at different airports.  Given this, we look to explore if operational efficiency within airlines affects the fares they are able to charge. 

```{r}
flight_data[is.na(flight_data)] <- 0
flight_data <- flight_data %>%
  filter(flight_data$airline_id %in% airline_ids_keep)

#Get the average delay for each domestic route in the US
airline_delay_by_airport <- flight_data %>%
  mutate(route = paste(origin_airport, destination_airport)) %>%
  group_by(airline_id, route) %>%
  summarize(delay_minutes = mean(airline_delay))

airline_delay_by_airport[1:2, ]
```

```{r}
#Join delay information with fare price information
fares_airline_caused_delays <- inner_join(airline_delay_by_airport, route_fares, by = c("airline_id", "route"))

fares_airline_caused_delays[1,]
```

```{r}
#Limit delays to 15 minutes as significant outliers skew the results
fares_airline_caused_delays <- fares_airline_caused_delays[!is.null(fares_airline_caused_delays)] %>%
  filter(delay_minutes < 15) #outliers

unique_airlines <- unique(fares_airline_caused_delays$airline_id)

#For each airline, produce a scatterplot relating fare prices to airline delays
for(airline in unique_airlines)
{
  selected_airline <- filter(fares_airline_caused_delays, airline_id == airline)
  
  plot(selected_airline$delay_minutes, selected_airline$avg_fare, xlab="Delay", ylab="Fare", sub=airline)
  
  abline(lm(selected_airline$avg_fare ~ selected_airline$delay_minutes))
  
  dev.copy(png,filename=paste(airline, ".png"))
  dev.off ()
}
```

